Equivalence Testing and the Second Generation \emph{P}-Value

### 2019-04-24 reply_to_reviews.R2.tex

\linenumbers

\authornote{

Correspondence concerning this article should be addressed to Daniël Lakens, Den Dolech 1, IPO 1.33, 5600 MB, Eindhoven, The Netherlands. E-mail: \href{mailto:D.Lakens@tue.nl}{\nolinkurl{D.Lakens@tue.nl}}}

\abstract{
In this second revision we are responding to the last issue you raised concerning our incorrect statement that 50\% of the expected correlations fall below the observed correlation. We are extremely grateful you pointed out this incorrect statement, and have corrected it as explained below. We have also made some minor changes after carefully re-reading the manuscript that we detail at the end of this letter. We hope you think our submission is ready to be accepted for publication in Meta-Psychology.

@@

In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to \emph{r} = -0.99 and 0.99, with a true correlation of 0.99) 97.60\% of the confidence interval overlaps with the equivalence range, even though in the long run only 36\% of the correlations observed in the future will fall in this range.

\begin{center}\rule{0.5\linewidth}{\linethickness}\end{center}

\textbf{We again want to sincerely thank you for catching this error before our paper was publlished.}

\hypertarget{additional-minor-changes}{%
\section{Additional Minor changes}\label{additional-minor-changes}}

\textbf{We replaced some values in the document that were hard-coded with the R scripts that produced these values, making the document (as far as we can see) 100\% reproducible. (note that this is not noticeable in the PDF)}

\textbf{We wrote we \enquote{replicated} Blume, but we now say we \enquote{build on} Blume - all our examples use a mean of 145, where upon rereading Blume et al we noticed they use a mean of 146 - an inconsequential difference, but instead of changing all our examples to a mean of 146, we just changed the word \enquote{replicated} to \enquote{build on}.}

\textbf{We removed \enquote{if too few observations are collected} which was redundant with the \enquote{note that when sample sizes are small} earlier in the same sentence.}

\textbf{We added in Figure 6 that the standard deviation for the visualized example is 2.}

\textbf{We added some clarification when describing Figure 12, replacing} \enquote{the two second generation \emph{p}-values associated with the observed correlations at \emph{r} = -0.45 and \emph{r} = 0.45 are larger than 50\%} \textbf{with} \enquote{the proportion of the confidence interval that overlaps with the equivalence range is larger than 50\% when the observed correlations are \emph{r} = -.45 and \emph{r} = .45, meaning that the two second generation p-values associated with these correlations are larger than 50\%.}

\textbf{In the conclusion, we noted that p-values from TOST are more consistent, and easier to interpret. Upon re-reading our manuscript, we chose to delete the \enquote{and easier to interpret} part of that sentence, since this is an empirical claim, and we have no data to support it, and is not relevant for the points we make.}

### 2019-04-24 manuscript.R2.tex
R2 submission

\documentclass[,man,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Equivalence Testing and the Second Generation P-Value},
            pdfauthor={Daniël Lakens~\& Marie Delacre},
            pdfkeywords={equivalence testing, second generation p-values, hypothesis testing, TOST, statistical inference},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\usepackage{lineno}

\linenumbers

@@

The second generation \emph{p}-value (SGPV) is calculated as:
\[
  p _ { \delta } = \frac { \left| I \cap H _ { 0 } \right| } { | I | } \times \max \left\{ \frac { | I | } { 2 \left| H _ { 0 } \right| } , 1 \right\}
\]
where I is the interval based on the data (e.g., a 95\% confidence interval) and H\textsubscript{0} is the equivalence range. The first term of this formula implies that the second generation \emph{p}-value is the width of the confidence interval that overlaps with the equivalence range, divided by the total width of the confidence interval. The second term is a \enquote{small sample correction} (which will be discussed later) that comes into play whenever the confidence interval is more than twice as wide as the equivalence range.
To examine the relation between the TOST \emph{p}-value and the SGPV we can calculate both statistics across a range of observed effect sizes. Replicating the example by Blume et al. (2018), in Figure \ref{fig:TOSTSGPV1} \emph{p}-values are plotted for the TOST procedure and the SGPV. The statistics are calculated for hypothetical one-sample \emph{t}-tests for observed means ranging from 140 to 150 (on the x-axis). The equivalence range is set to 145 ± 2 (i.e., an equivalence range from 143 to 147), the observed standard deviation is assumed to be 2, and the sample size is 30. For example, for the left-most point in Figure \ref{fig:TOSTSGPV1} the SGPV and the TOST \emph{p}-value is calculated for a hypothetical study with a sample size of 30, an observed standard deviation of 2, and an observed mean of 140, where the \emph{p}-value for the equivalence test is 1, and the SGPV is 0.

@@

\begin{figure}
\centering
\includegraphics{manuscript.R2_files/figure-latex/TOSTSGPV6-1.pdf}
\caption{\label{fig:TOSTSGPV6}Comparison of \emph{p}-values from TOST (black line) and SGPV (grey line) across a range of observed sample means (x-axis). Because the sample size is small (n = 10) and the CI is more than twice as wide as the equivalence range (set to -0.4 to 0.4), the SGPV is set to 0.5 (horizontal lightgrey line) across a range of observed means.}
\end{figure}

@@

Because the z-distribution is symmetric, the probability of observing the observed or more extreme z-score, assuming the equivalence bound is the true effect size, is 50\%. However, because the \emph{r} distribution is not symmetric, this does not mean that there is always a 50\% probability of observing a correlation smaller or larger than the true correlation. As can be seen in Figure \ref{fig:TOSTSGPV12}, the two second generation \emph{p}-values associated with the observed correlations at \emph{r} = -0.45 and \emph{r} = 0.45 are larger than 50\%. Because the confidence intervals are asymmetric around the observed effect size of 0.45 (ranging from 0.11 to 0.70) according to Blume et al. (2018) 58.11\% of the data-supported hypotheses are null hypotheses, and therefore 58.11\% of the data-supported hypotheses are compatible with the null premise.

@@

The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95\% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to \emph{r} = -0.99 and 0.99, with a true correlation of 0.99) 97.60\% of the confidence interval overlaps with the equivalence range, even though in the long run only 35.08\% of the correlations observed in the future will fall in this range.

@@

When introducing a new statistical method, it is important to compare it to existing approaches and specify its relative strengths and weaknesses. Here, we aimed to compare the SGPV against equivalence tests based on the TOST procedure. First of all, even though a SGPV of 1 or 0 has a clear interpretation (we can reject effects outside or inside the equivalence range), intermediate values are not as easy to interpret (especially for effects that have asymmetric confidence intervals). In one sense, they are what they are (the proportion of overlap), but it can be unclear what this number tells us about the data we have collected. This is not too problematic, since the main use of the SGPV (e.g., in all examples provided by Blume and colleagues) seems to be to examine whether the SGPV is 0, 1, or inconclusive. As already mentioned, this interpretation of a SGPV is very similar to the Neyman-Pearson interpretation of an equivalence test and a minimum effect tests (which are complementary). The difference is that where a SGPV of 1 can be interpreted as \emph{p} \textless{} .025, equivalence tests provide exact \emph{p}-values, and they continue to differentiate between for example \emph{p} = 0.024 and \emph{p} = 0.002. Whether this is desirable depends on the perspective that is used. From a Neyman-Pearson perspective on statistical inferences the main conclusion is based on whether or not \(p < \alpha\), and thus an equivalence test and SGPV can be performed by simply checking whether the confidence interval falls within the equivalence range, just as a null-hypothesis test can be performed by checking whether the confidence interval contains zero or not. At the same time, it is recommended to report exact \emph{p}-values (American Psychological Association, 2010), and exact \emph{p}-values might provide information of interest to readers about how precisely how surprising the data, or more extreme data, is under the null model. Some researchers might be interested in combining an equivalence test with a null-hypothesis significance test. This allows a researcher to ask whether there is an effect that is statistically different from zero, and whether effect sizes that are considered meaningful can be rejected. Equivalence tests combined with null-hypothesis tests classify results into four possible categories, and for example allow researchers to conclude an effect is significant \emph{and} equivalent (i.e., statistically different from zero, but also too small to be considered meaningful; see Lakens et al. (2018)).

@@

We believe that our explanation of the similarities between the TOST procedure and the SGPV provides context to interpret the contribution of second generation \emph{p}-values to the statistical toolbox. The novelty of the SGPV can be limited when confidence intervals are asymmetrical or wider than the equivalence range. There are strong similarities with \emph{p}-values from the TOST procedure, and in all situations where the statistics yield different results, the behavior of the \emph{p}-value from the TOST procedure is more consistent and easier to interpret. We hope this overview of the relationship between the SGPV and equivalence tests will help researchers to make an informed decision about which statistical approach provides the best answer to their question. Our comparisons show that when proposing alternatives to null-hypothesis tests, it is important to compare new proposals to already existing procedures. We believe equivalence tests achieve the goals of the second generation \emph{p}-value while allowing users to easily control error rates, and while yielding more consistent statistical outcomes.

@@

\leavevmode\hypertarget{ref-kruschke_rejecting_2018}{}%
Kruschke, J. K. (2018). Rejecting or Accepting Parameter Values in Bayesian Estimation. \emph{Advances in Methods and Practices in Psychological Science}, 2515245918771304. doi:\href{https://doi.org/10.1177/2515245918771304}{10.1177/2515245918771304}

@@

\leavevmode\hypertarget{ref-lakens_equivalence_2018}{}%
Lakens, D., Scheel, A. M., \& Isager, P. M. (2018). Equivalence Testing for Psychological Research: A Tutorial. \emph{Advances in Methods and Practices in Psychological Science}, 2515245918770963. doi:\href{https://doi.org/10.1177/2515245918770963}{10.1177/2515245918770963}

\leavevmode\hypertarget{ref-meyners_equivalence_2012}{}%
Meyners, M. (2012). Equivalence tests review. \emph{Food Quality and Preference}, \emph{26}(2), 231--245. doi:\href{https://doi.org/10.1016/j.foodqual.2012.05.003}{10.1016/j.foodqual.2012.05.003}

@@

\leavevmode\hypertarget{ref-schuirmann_comparison_1987}{}%
Schuirmann, D. J. (1987). A comparison of the two one-sided tests procedure and the power approach for assessing the equivalence of average bioavailability. \emph{Journal of Pharmacokinetics and Biopharmaceutics}, \emph{15}(6), 657--680.

\leavevmode\hypertarget{ref-serlin_rationality_1985}{}%
Serlin, R. C., \& Lapsley, D. K. (1985). Rationality in psychological research: The good-enough principle.

>>>>

\documentclass[,jou,floatsintext]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Equivalence Testing and the Second Generation P-Value},
            pdfauthor={Daniël Lakens~\& Marie Delacre},
            pdfkeywords={equivalence testing, second generation p-values, hypothesis testing, TOST, statistical inference},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

@@

The second generation \emph{p}-value (SGPV) is calculated as:
\[
  p _ { \delta } = \frac { \left| I \cap H _ { 0 } \right| } { | I | } \times \max \left\{ \frac { | I | } { 2 \left| H _ { 0 } \right| } , 1 \right\}
\]
where I is the interval based on the data (e.g., a 95\% confidence interval) and H\textsubscript{0} is the equivalence range. The first term of this formula implies that the second generation \emph{p}-value is the width of the confidence interval that overlaps with the equivalence range, divided by the total width of the confidence interval. The second term is a \enquote{small sample correction} (which will be discussed later) that comes into play whenever the confidence interval is more than twice as wide as the equivalence range.
To examine the relation between the TOST \emph{p}-value and the SGPV we can calculate both statistics across a range of observed effect sizes. Building on the example by Blume et al. (2018), in Figure \ref{fig:TOSTSGPV1} \emph{p}-values are plotted for the TOST procedure and the SGPV. The statistics are calculated for hypothetical one-sample \emph{t}-tests for observed means ranging from 140 to 150 (on the x-axis). The equivalence range is set to 145 ± 2 (i.e., an equivalence range from 143 to 147), the observed standard deviation is assumed to be 2, and the sample size is 30. For example, for the left-most point in Figure \ref{fig:TOSTSGPV1} the SGPV and the TOST \emph{p}-value is calculated for a hypothetical study with a sample size of 30, an observed standard deviation of 2, and an observed mean of 140, where the \emph{p}-value for the equivalence test is 1, and the SGPV is 0.

@@

\begin{figure}
\centering
\includegraphics{manuscript.R2_files/figure-latex/TOSTSGPV6-1.pdf}
\caption{\label{fig:TOSTSGPV6}Comparison of \emph{p}-values from TOST (black line) and SGPV (grey line) across a range of observed sample means (x-axis). Because the sample size is small (n = 10) and with a standard deviation of 2 the CI is more than twice as wide as the equivalence range (set to -0.4 to 0.4), the SGPV is set to 0.5 (horizontal lightgrey line) across a range of observed means.}
\end{figure}

@@

Because the z-distribution is symmetric, the probability of observing the observed or more extreme z-score, assuming the equivalence bound is the true effect size, is 50\%. However, because the \emph{r} distribution is not symmetric, this does not mean that there is always a 50\% probability of observing a correlation smaller or larger than the true correlation. As can be seen in Figure \ref{fig:TOSTSGPV12}, the proportion of the confidence interval that overlaps with the equivalence range is larger than 50\% when the observed correlations are \emph{r} = -.45 and \emph{r} = .45, meaning that the two second generation p-values associated with these correlations are larger than 50\%. Because the confidence intervals are asymmetric around the observed effect size of 0.45 (ranging from 0.11 to 0.70) according to Blume et al. (2018) 58.11\% of the data-supported hypotheses are null hypotheses, and therefore 58.11\% of the data-supported hypotheses are compatible with the null premise.

@@

The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95\% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to \emph{r} = -0.99 and 0.99, with a true correlation of 0.99) 97.60\% of the confidence interval overlaps with the equivalence range, even though in the long run only 36\% of the correlations observed in the future will fall in this range.

@@

When introducing a new statistical method, it is important to compare it to existing approaches and specify its relative strengths and weaknesses. Here, we aimed to compare the SGPV against equivalence tests based on the TOST procedure. First of all, even though a SGPV of 1 or 0 has a clear interpretation (we can reject effects outside or inside the equivalence range), intermediate values are not as easy to interpret (especially for effects that have asymmetric confidence intervals). In one sense, they are what they are (the proportion of overlap), but it can be unclear what this number tells us about the data we have collected. This is not too problematic, since the main use of the SGPV (e.g., in all examples provided by Blume and colleagues) seems to be to examine whether the SGPV is 0, 1, or inconclusive. As already mentioned, this interpretation of a SGPV is very similar to the Neyman-Pearson interpretation of an equivalence test and a minimum effect tests (which are complementary). The difference is that where a SGPV of 1 can be interpreted as \emph{p} \textless{} .025, equivalence tests provide exact \emph{p}-values, and they continue to differentiate between for example \emph{p} = 0.024 and \emph{p} = 0.002. Whether this is desirable depends on the perspective that is used. From a Neyman-Pearson perspective on statistical inferences the main conclusion is based on whether or not \(p < \alpha\), and thus an equivalence test and SGPV can be performed by simply checking whether the confidence interval falls within the equivalence range, just as a null-hypothesis test can be performed by checking whether the confidence interval contains zero or not. At the same time, it is recommended to report exact \emph{p}-values (American Psychological Association, 2010), and exact \emph{p}-values might provide information of interest to readers about how precisely how surprising the data, or more extreme data, is under the null model. Some researchers might be interested in combining an equivalence test with a null-hypothesis significance test. This allows a researcher to ask whether there is an effect that is statistically different from zero, and whether effect sizes that are considered meaningful can be rejected. Equivalence tests combined with null-hypothesis tests classify results into four possible categories, and for example allow researchers to conclude an effect is significant \emph{and} equivalent (i.e., statistically different from zero, but also too small to be considered meaningful; see Lakens et al., 2018).

@@

We believe that our explanation of the similarities between the TOST procedure and the SGPV provides context to interpret the contribution of second generation \emph{p}-values to the statistical toolbox. The novelty of the SGPV can be limited when confidence intervals are asymmetrical or wider than the equivalence range. There are strong similarities with \emph{p}-values from the TOST procedure, and in all situations where the statistics yield different results, the behavior of the \emph{p}-value from the TOST procedure is more consistent. We hope this overview of the relationship between the SGPV and equivalence tests will help researchers to make an informed decision about which statistical approach provides the best answer to their question. Our comparisons show that when proposing alternatives to null-hypothesis tests, it is important to compare new proposals to already existing procedures. We believe equivalence tests achieve the goals of the second generation \emph{p}-value while allowing users to easily control error rates, and while yielding more consistent statistical outcomes.

@@

\leavevmode\hypertarget{ref-kruschke_rejecting_2018}{}%
Kruschke, J. K. (2018). Rejecting or Accepting Parameter Values in Bayesian Estimation. \emph{Advances in Methods and Practices in Psychological Science}, \emph{1}(2), 270--280. doi:\href{https://doi.org/10.1177/2515245918771304}{10.1177/2515245918771304}

@@

\leavevmode\hypertarget{ref-lakens_equivalence_2018-1}{}%
Lakens, D., Scheel, A. M., \& Isager, P. M. (2018). Equivalence Testing for Psychological Research: A Tutorial. \emph{Advances in Methods and Practices in Psychological Science}, \emph{1}(2), 259--269. doi:\href{https://doi.org/10.1177/2515245918770963}{10.1177/2515245918770963}

\leavevmode\hypertarget{ref-meyners_equivalence_2012}{}%
Meyners, M. (2012). Equivalence tests A review. \emph{Food Quality and Preference}, \emph{26}(2), 231--245. doi:\href{https://doi.org/10.1016/j.foodqual.2012.05.003}{10.1016/j.foodqual.2012.05.003}

@@

\leavevmode\hypertarget{ref-schuirmann_comparison_1987}{}%
Schuirmann, D. J. (1987). A comparison of the two one-sided tests procedure and the power approach for assessing the equivalence of average bioavailability. \emph{Journal of Pharmacokinetics and Biopharmaceutics}, \emph{15}(6), 657--680. doi:\href{https://doi.org/https://doi.org/10.1007/BF01068419}{https://doi.org/10.1007/BF01068419}

\leavevmode\hypertarget{ref-serlin_rationality_1985}{}%
Serlin, R. C., \& Lapsley, D. K. (1985). Rationality in psychological research: The good-enough principle. \emph{American Psychologist}, \emph{40}(1), 73--83. doi:\href{https://doi.org/http://dx.doi.org/10.1037/0003-066X.40.1.73}{http://dx.doi.org/10.1037/0003-066X.40.1.73}
